Pollution, agricultural productivity, and development: Evidence from coal plants in India^[I thank Marc Bellemare and David Sungho Park for comments and suggestions that greatly improved the quality of the paper. The usual disclaimer applies. All scripts used in this paper are publicly available on 

### 2023-03-02 draft/draft.Rmd
draft update

To identify the effects of air pollution on agricultural productivity, I instead combine the roll out of coal plants in India with plausibly exogenous changes in wind direction -- in a similar spirit to @deryugina2019mortality -- to measure the effects of pollution from coal plants on key development outcomes. I first plot the location of coal plants in India by year. Using a village-level shapefile, I identify all villages that are located within 30km of a coal plant at any time during the sample period, which comprises the 1990s to the 2010s, with differences depending on data availability.

@@

This paper contributes to several strands of literature. First, we already have evidence of effects of pollution on different forms of labor activity. For example, general levels of pollution decrease productivity of call center works in China [@chang2019effect], lead people to perform worse on cognitive functions [@ebenstein2016long; @wen2022lower], and also lead to lower levels of farm labor productivity in California, driven by changes in ozone concentration [@graff2012impact]. Also in China, @he2019severe find small but significant negative effects of pollution on manufacturing productivity, while @chen2022effect show pollution leads to labor out migration. Perhaps relevant to my labor allocation results, @hanna2015effect show that decreases in pollution lead to increases in labor allocation, while I show the same relationship among the oldest cohort. The papers most related to this one are @aragon2016polluting, who show that gold mines in Ghana have substantial negative impacts on agricultural productivity in the surrounding areas, and @burney2014recent, who model the effects of climate change and pollutants on agricultural productivity in India. This paper differs in two important ways. First, I explore effects of pollution from coal plants, which are much more ubiquitous than gold mines across the world. Second, I use a different identification strategy from either paper, relying on wind blowing pollution from coal plants.

@@

The main goal of this paper is to examine whether exposure to emissions from coal plants affects agricultural productivity, labor allocation, and economic growth. I use several sources of data, which I describe here and in Table \@ref(tab:data) in the appendix.

@@

The second dataset includes agricultural productivity for both the monsoon and winter seasons, from 2002 to 2018. To match other data, I only use the data up to 2013. This data comes from @gangopadhyay2022new and estimates land productivity (i.e. yield, in tons per hectare) for the major crops in India, using satellite-derived data. While there is a wealth of literature on measurement error in land [@desiere2018land; @abay2022nonclassical], the authors use a static cropland mask for the productivity estimates. This means that any measurement error will be constant across time and subsumed by the village fixed effects. While the measure of output is also noisy, the use of satellite-derived data and a static cropland mask reduces concerns about measurement error being responsible for the results below. The authors define the monsoon season as June to October and the winter season as November to March. I keep these definitions when matching across data below. This data is also publicly available from Nature's data-sharing website.[^7] Since the main regressions below use village fixed effects, any differences in cropping patterns across areas should not bias estimates on changes, as long as these patterns do not change markedly in response to pollution.

@@

The agricultural productivity data is available as raster files with a resolution of 500m. To aggregate this data up to a useful administrative unit, I use village-level shapefiles provided by @almn2021 and publicly available on the SHRUG platform.[^8] I aggregate the agricultural productivity data to the village level by extracting mean productivity for each feature in the shapefile. I do this separately for each season -- monsoon and winter -- and each year, from 2002 to 2013.

@@

To measure exposure to pollution from coal plants, I first locate all village centroids located within 30km from a coal plant in a given year. I choose 30km due to previous research on the effects of (air) pollution [@aragon2016polluting; @bencko1977health; @li2014health]. I then calculate the direction from coal plants to all village centroids within that 30km radius. To define exposure, I then pull daily wind direction data from the National Center for Atmospheric Research.[^9] For each day, I document whether the wind is blowing towards each village centroid.[^10] I then temporally aggregate this daily data depending on the temporal definition of the corresponding outcome. For example, agricultural productivity is defined across five months (e.g. the monsoon season is from June to October) so I count the total days a given village is exposed to wind blowing from *any* coal plant within 30km during those five months.

@@

I am primarily interested in the effects of pollution on agricultural productivity. We might consider a regression of the following form:

\begin{gather} y_{it} = \alpha_{i} + \gamma_{t} + \beta PM_{it} + \delta rain_{it} + \psi temp_{it} + \varepsilon_{it}, \end{gather}

where $y_{it}$ is outcome $y$ for unit $i$ at time $t$, $\alpha$ is a geographic fixed effects, $\gamma_t$ is time fixed effects, $PM_{it}$ is the concentration of particulate matter 2.5, $rain_{it}$ is total rainfall during the season, and $temp_{it}$ is average temperature during the season.

However, even with the fixed effects, this regression likely suffers from endogeneity; places with more pollution may be very different from places with less pollution. It is difficult to sign the direction of the bias a priori, because it depends on how the coal plants are placed. For example, if coal plants are placed in places in need of power -- for example, due to rapid growth -- then the estimates might be upwards biased. On the other hand, if official purposefully place coal plants in poorer areas for political reasons, then the estimates might be downwards biased.

So what are the characteristics of areas near the construction of new coal plants? Table \@ref(tab:plantresultstable) presents four separate regressions. The first two columns use 1991 census data -- downloaded from SHRUG -- and the second two columns use 2001 census data, all at the village level. The first column presents results from regressing a dummy for whether there is a coal plant within 30km in 1991 on several census variables. Villages with higher populations and higher literacy in 1991 are more likely to be near a coal plant at the time. Interestingly, however, they are less likely to have a paved road.

@@

The second column keeps the 1991 census data but changes the outcome variable to a dummy for whether there *will be* a coal plant constructed within 30km before 2001. This column restricts the sample to only villages that do not already have a coal plant within 30km in 1991. Population is no longer a significant predictor while the coefficients on literacy and the presence of a paved road have flipped.

@@

Identification relies on changes in wind direction across time for the same geographic units. In other words, conditional on the fixed effects and weather, I assume changes in wind direction are as good as random and uncorrelated with the outcome -- e.g. agricultural productivity -- except through exposure to pollution. One major threat to identification is that the building of a coal plant could lead to a spurious correlation between the exposure variable and the outcome. Specifically, since exposure is defined as zero prior to the building of a coal plant, the exposure variable itself might be endogenous even though wind direction is not. To help allay these concerns, I estimate additional regressions restricting the sample to only time periods after which a given location has a coal plant. I also present results using leads to show that future wind direction does not seem to predict current outcomes.

@@

In addition to serving as an example, Table \@ref(tab:pollutiontable) also helps validate the use of wind direction as a proxy for exposure to pollution. The first two columns include all available years of data while the last two columns restrict estimation only to the years for which the agricultural productivity regressions are estimated. Across all four columns, the story is the same: when the wind is blowing in the direction of a village from a coal plant within 30 kilometers, estimated $\mathrm{PM_{2.5}}$ is substantially higher. The regressions indicate that one additional day of wind in the direction of the village increases mean $\mathrm{PM_{2.5}}$ for the month by between `r paste0(substr(pollutiontable[1,2],1,5))` and `r paste0(substr(pollutiontable[1,3],1,5))` $\mathrm{\mu g/m^3}$.

@@

Consider the following regression of interest:
\begin{gather} 
y_{it} = \alpha_{i} + \gamma_{t} + \omega pm_{it} + \phi open_{it} + \beta pm_{it}\times open_{it} + \delta rain_{it} + \psi temp_{it} + \varepsilon_{it},
\end{gather}

where the interaction term $pm_{it}\times open_{it}$ indicates the change in the effect of pollution following the opening of the coal plant. Note that this is not the same as saying that pollution increased; instead, it is focusing on the change in the effect of a change in pollution of one unit on the outcome of interest, agricultural yield. I instrument for both $pm_{it}$ and $pm_{it}\times open_{it}$ using wind and the interaction between wind and the open indicator, or $wind_{it}\times open_{it}$.

@@

I first present results for agricultural productivity, with the reduced form. The outcome of all the regressions in this section is the log of agricultural land productivity, which is defined as tons per hectare. The first set of results are in Table \@ref(tab:yieldtable). The unit of analysis is the village-year -- from 2002 to 2013 -- focusing only on the monsoon (kharif) season. The data used to construct the yield measure is calculated using five months of data, so the wind variable is defined for the same five months and is the sum of days in which the wind is pointed towards the village centroid from a coal plant.

@@

Of course, it is not wind, per se, that we are interested in. Instead, we are actually interested in better understanding the effects of exposure to pollution on agricultural productivity. Table \@ref(tab:yieldtablepm) presents the same regressions as Table \@ref(fig:yieldtable), with a key difference: instead of wind exposure, it uses logged average concentration of $\mathrm{PM_{2.5}}$ -- from @hammer2020global -- during the season as the key explanatory variable. There is again a consistent, negative relationship between particulate matter and agricultural productivity. Since both variables are logged, the coefficients are similar to elasticities, indicating that a one-percent increase in pollution leads to a 0.03-to-0.04 decrease in agricultural productivity.

Recall, however, the results from above: coal plants were placed in places with higher average agricultural productivity. This means that the simple regressions in \@ref(tab:yieldtablepm) might actually be biased upwards: although pollution might negatively affect productivity, pollution is higher in more productive areas due to the endogenous placement of coal plants in the study areas. To recover the actual effect of pollution, we need an instrument. If, conditional on the fixed effects and weather variables, wind direction is exogenous and only affects agricultural productivity through pollution, then it is a valid instrument for pollution.

I present the IV results in Table \@ref(tab:yieldtabletwo), with the first-stage F-statistic in the bottom. The first-stage results make clear how strong of an instrument wind is; all three F-statistics are above 1,000, meeting even the higher bar for the first-stage F-test argued for by @lee2022valid. Also note that the coefficients are quite different from those in Table \@ref(tab:pollutiontable) because the unit of analysis in Table \@ref(tab:yieldtabletwo) is the season, while in Table \@ref(tab:pollutiontable) the unit of analysis is the month.

@@

Having established that exposure to pollution has a negative effect on agricultural productivity, I now move to heterogeneity in these effects. Table \@ref(tab:yieldtablehet) presents three different analyses: heterogeneity by median exposure, heterogeneity by initial yield, and possible multiplicative effects of shocks. All regressions use the IV strategy detailed above and I include the F statistic from the first stage regressions (of which there are two in the last column) at the bottom of the table. Table \@ref(tab:yieldtablehet1st) in the appendix presents the first-stage results for all columns.

>>>>

To identify the effects of air pollution on agricultural productivity, I instead combine the roll out of coal plants in India with plausibly exogenous changes in wind direction -- in a similar spirit to @deryugina2019mortality -- to measure the effects of pollution from coal plants on key development outcomes. I first plot the location of coal plants in India by year. Using a village-level shapefile, I identify all villages that are located within 30km of a coal plant at any time during the sample period, which comprises the 2000s to the 2010s, with differences across outcomes depending on data availability.

@@

This paper contributes to several strands of literature. First, we already have evidence of effects of pollution on different forms of labor activity. For example, general levels of pollution decrease productivity of call center works in China [@chang2019effect], lead people to perform worse on cognitive functions [@ebenstein2016long; @wen2022lower], and also lead to lower levels of farm labor productivity in California, driven by changes in ozone concentration [@graff2012impact]. Also in China, @he2019severe find small but significant negative effects of pollution on manufacturing productivity, while @chen2022effect show pollution leads to labor out migration. Perhaps relevant to my labor allocation results, @hanna2015effect show that decreases in pollution lead to increases in labor allocation, while I show the same relationship in at least a subset of months of the agricultural season. The papers most related to this one are @aragon2016polluting, who show that gold mines in Ghana have substantial negative impacts on agricultural productivity in the surrounding areas, and @burney2014recent, who model the effects of climate change and pollutants on agricultural productivity in India. This paper differs in two important ways. First, I explore effects of pollution from coal plants, which are much more ubiquitous than gold mines across the world. Second, I use a different identification strategy from either paper, relying on wind blowing pollution from coal plants.

@@

The main goal of this paper is to examine whether exposure to pollution and emissions from coal plants, in particular, affect agricultural productivity. I use several sources of data, which I describe here and list in Table \@ref(tab:data) of the appendix.

@@

The second dataset includes agricultural productivity for both the monsoon and winter seasons, from 2002 to 2018. To match other data, I only use the data up to 2013. This data comes from @gangopadhyay2022new and estimates land productivity (i.e. yield, in tons per hectare) for the major crops in India, using satellite-derived data. While there is a wealth of literature on measurement error in land [@desiere2018land; @abay2022nonclassical], the authors use a static cropland mask for the productivity estimates. This means that any measurement error will be constant across time and subsumed by the village fixed effects. While the measure of output is also noisy, the use of satellite-derived data and a static cropland mask reduces concerns about measurement error being responsible for the results below, especially since common sources of measurement error are unlikely to be correlated with wind direction. The authors define the monsoon season as June to October and the winter season as November to March. I focus on only the monsoon season and keep the authors' definition when matching across data below. This data is also publicly available from Nature's data-sharing website.[^7] Since the main regressions below use village fixed effects, any differences in cropping patterns across areas should not bias estimates on changes, as long as these patterns do not change markedly in response to pollution.

@@

The agricultural productivity data is available as raster files with a resolution of 500m. To aggregate this data up to a useful administrative unit, I use village-level shapefiles provided by @almn2021 and publicly available on the SHRUG platform.[^8] I aggregate the agricultural productivity data to the village level by extracting mean productivity for each feature in the shapefile. I do this separately for the monsoon season each year, from 2002 to 2013.

@@

To measure exposure to pollution from coal plants, I first locate all village centroids located within 30km from a coal plant in a given year. I choose 30km due to previous research on the effects of (air) pollution [@aragon2016polluting; @bencko1977health; @li2014health]. I then calculate the direction from coal plants to all village centroids within that 30km radius. To define exposure, I then pull daily wind direction data from the National Center for Atmospheric Research.[^9] For each day, I document whether the wind is blowing towards each village centroid[^10] and then temporally aggregate this daily data depending on the temporal definition of the corresponding outcome. For example, agricultural productivity is defined across five months (e.g. the monsoon season is from June to October) so I count the total days a given village is exposed to wind blowing from *any* coal plant within 30km during those five months.

@@

I am primarily interested in the effects of pollution on agricultural productivity. We might consider a regression of the following form:
\begin{gather} y_{it} = \alpha_{i} + \gamma_{t} + \beta PM_{it} + \delta rain_{it} + \psi temp_{it} + \varepsilon_{it}, \end{gather}

where $y_{it}$ is outcome $y$ for unit $i$ at time $t$, $\alpha$ is a geographic fixed effects, $\gamma_t$ is time fixed effects, $PM_{it}$ is the (log) concentration of particulate matter 2.5, $rain_{it}$ is total rainfall during the season, and $temp_{it}$ is average temperature during the season.

However, even with the fixed effects, this regression likely suffers from endogeneity; places with more pollution may be very different from places with less pollution. It is difficult to sign the direction of the bias a priori, because it depends on how the coal plants are placed. For example, if coal plants are placed in places in need of power -- for example, due to rapid growth -- then the estimates might be upwards biased. On the other hand, if officials purposefully place coal plants in poorer areas for political reasons, then the estimates might be downwards biased.

So what are the characteristics of areas near the construction of new coal plants? Table \@ref(tab:plantresultstable) presents four separate regressions. The first two columns use 1991 census data -- downloaded from SHRUG -- and the second two columns use 2001 census data, all at the village level. The first column presents results from regressing a dummy for whether there is a coal plant within 30km in 1991 on several census variables. Villages with higher populations and higher literacy in 1991 are more likely to be near a coal plant at the time.

@@

The second column keeps the 1991 census data but changes the outcome variable to a dummy for whether there *will be* a coal plant constructed within 30km before 2001. This column restricts the sample to only villages that do not already have a coal plant within 30km in 1991. Population and literacy both remain strong predictors of the building of a new coal plant.

@@

Identification relies on changes in wind direction across time for the same geographic units. In other words, conditional on the fixed effects and weather, I assume changes in wind direction are as good as random and uncorrelated with the outcome -- primarily agricultural productivity in this paper -- except through exposure to pollution.

@@

In addition to serving as an example, Table \@ref(tab:pollutiontable) also helps validate the use of wind direction as a proxy for exposure to pollution. The first two columns include all available years of data while the last two columns restrict estimation only to the years for which the agricultural productivity regressions are estimated. Across all four columns, the story is the same: when the wind is blowing in the direction of a village from a coal plant within 30 kilometers, estimated $\mathrm{PM_{2.5}}$ is substantially higher.

@@

Consider the following regression of interest:
\begin{gather} 
y_{it} = \alpha_{i} + \gamma_{t} + \omega PM_{it} + \phi open_{it} + \beta PM_{it}\times open_{it} + \delta rain_{it} + \psi temp_{it} + \varepsilon_{it},
\end{gather}

where the interaction term $PM_{it}\times open_{it}$ indicates the change in the effect of pollution following the opening of the coal plant. Note that this is not the same as saying that pollution increased; instead, it is focusing on the change in the effect of a change in pollution of one unit on the outcome of interest, agricultural yield. I instrument for both $PM_{it}$ and $PM_{it}\times open_{it}$ using wind and the interaction between wind and the open indicator, or $wind_{it}\times open_{it}$.

@@

I first present the reduced form results for agricultural productivity. The outcome of all the regressions in this section is the log of agricultural land productivity, which is defined as tons per hectare. The first set of results are in Table \@ref(tab:yieldtable). The unit of analysis is the village-year -- from 2002 to 2013 -- focusing only on the monsoon (kharif) season. The data used to construct the yield measure is calculated using five months of data, so the wind variable is defined for the same five months and is the sum of days in which the wind is pointed towards the village centroid from a coal plant.

@@

Of course, it is not wind, per se, that we are interested in. Instead, we are actually interested in better understanding the effects of exposure to pollution on agricultural productivity. Table \@ref(tab:yieldtablepm) presents the same regressions as Table \@ref(tab:yieldtable), with a key difference: instead of wind exposure, it uses log average concentration of $\mathrm{PM_{2.5}}$ -- from @hammer2020global -- during the season as the key explanatory variable. There is again a consistent, negative relationship between particulate matter and agricultural productivity. Since both variables are logged, the coefficients are similar to elasticities, indicating that a one-percent increase in pollution leads to a 0.03-to-0.04 percent decrease in agricultural productivity.

Recall, however, the results from above: coal plants were placed in places with higher average agricultural productivity. This means that the simple regressions in Table \@ref(tab:yieldtablepm) might actually be biased upwards: although pollution might negatively affect productivity, pollution is higher in more productive areas due to the endogenous placement of coal plants in the study areas. To recover the actual effect of pollution, we need an instrument. If, conditional on the fixed effects and weather variables, wind direction is exogenous and only affects agricultural productivity through pollution, then it is a valid instrument for pollution.

I present the IV results in Table \@ref(tab:yieldtabletwo), with the first-stage F-statistic at the bottom. The first-stage results make clear how strong of an instrument wind is; all three F-statistics are above 1,000, meeting even the higher bar for the first-stage F-test argued for by @lee2022valid. Also note that the coefficients are quite different from those in Table \@ref(tab:pollutiontable) because the unit of analysis in Table \@ref(tab:yieldtabletwo) is the season, while in Table \@ref(tab:pollutiontable) the unit of analysis is the month.

@@

Having established that exposure to pollution has a negative effect on agricultural productivity, I now move to heterogeneity in these effects. Table \@ref(tab:yieldtablehet) presents three different analyses: heterogeneity by median exposure, heterogeneity by initial yield, and possible multiplicative effects of shocks. All regressions use the IV strategy detailed above and I include the F statistic from the first stage regressions (of which there are two in the last column) at the bottom of the table.
